---
title: "Gotham Loops Sequential Geometry: Hypothesis Notebook"
output:
  word_document:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)

project_root <- if (
  file.exists("DESCRIPTION")
) {
  "."
} else if (
  file.exists("../DESCRIPTION")
) {
  ".."
} else {
  stop("Could not locate the gothamloops.stats package root.")
}

devtools::load_all(
  project_root,
  quiet = TRUE
)

exploration_script <- file.path(
  project_root,
  "data-raw",
  "sequence_grid_exploration.R"
)

if (!file.exists(exploration_script)) {
  stop(
    "Could not locate data-raw/sequence_grid_exploration.R. ",
    "Apply the sequential-grid exploration harness before knitting this report."
  )
}

source(
  exploration_script,
  local = knitr::knit_global()
)

stochastic_script <- file.path(
  project_root,
  "data-raw",
  "stochastic_alignment_exploration.R"
)

if (!file.exists(stochastic_script)) {
  stop(
    "Could not locate data-raw/stochastic_alignment_exploration.R. ",
    "Apply the stochastic-alignment exploration harness before knitting this report."
  )
}

source(
  stochastic_script,
  local = knitr::knit_global()
)

dat <- data.frame(
  group = rep(c("A", "B", "C"), each = 3),
  y = 1:9
)

state <- gotham_state(
  dat,
  outcome = "y",
  group = "group"
)

tau <- 0.80

alignment_grid <- seq(
  -1,
  1,
  by = 0.25
)

review <- run_gotham_sequence_grid_review(
  state = state,
  steps = 50,
  threshold = tau,
  alignment = alignment_grid
)

trajectory <- review$trajectory
diagnostics <- review$frontier_diagnostics
endpoint <- review$summary

zero_icc_formula <- function(n, rho) {
  (
    n^2 * rho +
      n * rho -
      2 * rho +
      2
  ) / (
    n^2 +
      n * rho -
      rho +
      1
  )
}

zero_icc_headroom_formula <- function(n, rho) {
  (
    (1 - rho) *
      (n * rho - rho + 2)
  ) / (
    n^2 +
      n * rho -
      rho +
      1
  )
}

infinity_icc_formula <- function(n, g) {
  1 / (
    g +
      n * (g - 1)
  )
}

normalized_frontier_radius <- function(
    n,
    g,
    tau,
    alignment) {
  d_n <- g + n * (g - 1)

  a_n <- (
    tau * d_n - 1
  ) / (
    g * (n + 1) * (1 - tau)
  )

  gamma_n <- (
    tau * (n - 1) + 2
  ) / (
    tau * (n - 1) + 1
  )

  if (any(a_n <= 0)) {
    stop(
      "The normalized frontier formula requires the H2 regime with a_n > 0."
    )
  }

  (
    alignment +
      sqrt(
        alignment^2 +
          a_n * gamma_n
      )
  ) / a_n
}

normalized_frontier_radius_limit <- function(
    g,
    tau,
    alignment) {
  a_infinity <- tau * (g - 1) /
    (
      g * (1 - tau)
    )

  (
    alignment +
      sqrt(
        alignment^2 +
          a_infinity
      )
  ) / a_infinity
}

asymptotic_alignment_drift <- function(
    g,
    tau,
    alignment) {
  2 * alignment *
    normalized_frontier_radius_limit(
      g = g,
      tau = tau,
      alignment = alignment
    )
}

run_normalized_recurrence <- function(
    initial_A,
    initial_n,
    alignment,
    g,
    tau) {
  steps <- length(alignment)

  out <- data.frame(
    step = 0:steps,
    group_size = initial_n + 0:steps,
    alignment = c(NA_real_, alignment),
    group_offset_ss = NA_real_,
    stringsAsFactors = FALSE
  )

  out$group_offset_ss[[1L]] <- initial_A

  if (steps == 0L) {
    return(out)
  }

  for (i in seq_len(steps)) {
    n_i <- initial_n + i - 1L
    c_i <- alignment[[i]]

    z_i <- normalized_frontier_radius(
      n = n_i,
      g = g,
      tau = tau,
      alignment = c_i
    )

    multiplier_i <- 1 +
      2 * c_i * z_i / (n_i + 1) +
      z_i^2 / (n_i + 1)^2

    out$group_offset_ss[[i + 1L]] <-
      out$group_offset_ss[[i]] *
      multiplier_i
  }

  out
}

normalized_multiplier <- function(
    n,
    g,
    tau,
    alignment) {
  z <- normalized_frontier_radius(
    n = n,
    g = g,
    tau = tau,
    alignment = alignment
  )

  1 +
    2 * alignment * z / (n + 1) +
    z^2 / (n + 1)^2
}

neutral_positive_share_asymptotic <- function(
    g,
    tau,
    negative_alignment,
    positive_alignment) {
  if (
    negative_alignment >= 0 ||
      positive_alignment <= 0
  ) {
    stop(
      "Require negative_alignment < 0 < positive_alignment."
    )
  }

  h_negative <- asymptotic_alignment_drift(
    g = g,
    tau = tau,
    alignment = negative_alignment
  )

  h_positive <- asymptotic_alignment_drift(
    g = g,
    tau = tau,
    alignment = positive_alignment
  )

  -h_negative /
    (
      h_positive -
        h_negative
    )
}

neutral_positive_share_finite_n <- function(
    n,
    g,
    tau,
    negative_alignment,
    positive_alignment) {
  m_negative <- normalized_multiplier(
    n = n,
    g = g,
    tau = tau,
    alignment = negative_alignment
  )

  m_positive <- normalized_multiplier(
    n = n,
    g = g,
    tau = tau,
    alignment = positive_alignment
  )

  -log(m_negative) /
    (
      log(m_positive) -
        log(m_negative)
    )
}

exact_structural_neutral_alignment <- function(
    n,
    g,
    tau) {
  numerator <-
    g *
    (1 - tau) *
    (
      tau * (n - 1) + 2
    )

  denominator <-
    4 *
    (n + 1) *
    (g - 1) *
    (1 + n * tau) *
    (
      tau * (n - 1) + 1
    )

  -sqrt(
    numerator /
      denominator
  )
}

exact_structural_neutral_normalized_radius <- function(
    n,
    g,
    tau) {
  c_star <- exact_structural_neutral_alignment(
    n = n,
    g = g,
    tau = tau
  )

  -2 *
    c_star *
    (n + 1)
}

run_greedy_neutral_controller <- function(
    initial_A,
    initial_n,
    steps,
    g,
    tau,
    negative_alignment,
    positive_alignment) {
  out <- data.frame(
    step = 0:steps,
    group_size = initial_n + 0:steps,
    alignment = NA_real_,
    group_offset_ss = NA_real_,
    log_ratio_to_initial = NA_real_,
    stringsAsFactors = FALSE
  )

  out$group_offset_ss[[1L]] <- initial_A
  out$log_ratio_to_initial[[1L]] <- 0

  cumulative_log_ratio <- 0

  for (i in seq_len(steps)) {
    n_i <- initial_n + i - 1L

    m_negative <- normalized_multiplier(
      n = n_i,
      g = g,
      tau = tau,
      alignment = negative_alignment
    )

    m_positive <- normalized_multiplier(
      n = n_i,
      g = g,
      tau = tau,
      alignment = positive_alignment
    )

    candidate_negative <-
      cumulative_log_ratio +
      log(m_negative)

    candidate_positive <-
      cumulative_log_ratio +
      log(m_positive)

    if (
      abs(candidate_positive) <
        abs(candidate_negative)
    ) {
      chosen_alignment <- positive_alignment
      chosen_multiplier <- m_positive
      cumulative_log_ratio <- candidate_positive
    } else {
      chosen_alignment <- negative_alignment
      chosen_multiplier <- m_negative
      cumulative_log_ratio <- candidate_negative
    }

    out$alignment[[i + 1L]] <-
      chosen_alignment

    out$group_offset_ss[[i + 1L]] <-
      out$group_offset_ss[[i]] *
      chosen_multiplier

    out$log_ratio_to_initial[[i + 1L]] <-
      cumulative_log_ratio
  }

  out
}

diagnostics$icc_at_zero_formula <- zero_icc_formula(
  diagnostics$group_size,
  diagnostics$icc_before
)

diagnostics$icc_at_infinity_formula <- infinity_icc_formula(
  diagnostics$group_size,
  state$n_groups
)

zero_formula_max_error <- max(
  abs(
    diagnostics$icc_at_zero -
      diagnostics$icc_at_zero_formula
  )
)

threshold_rows <- diagnostics$step > 1L

zero_threshold_formula_max_error <- max(
  abs(
    diagnostics$icc_at_zero[threshold_rows] -
      zero_icc_formula(
        diagnostics$group_size[threshold_rows],
        tau
      )
  )
)

infinity_formula_max_error <- max(
  abs(
    diagnostics$icc_at_infinity -
      diagnostics$icc_at_infinity_formula
  )
)

zero_steps <- sort(
  unique(
    diagnostics$step
  )
)

zero_spread <- do.call(
  rbind,
  lapply(
    zero_steps,
    function(step_value) {
      x <- diagnostics$icc_at_zero[
        diagnostics$step == step_value
      ]

      data.frame(
        step = step_value,
        min = min(x),
        max = max(x),
        spread = max(x) - min(x)
      )
    }
  )
)

root_counts <- as.data.frame(
  table(
    diagnostics$n_roots
  )
)

names(root_counts) <- c(
  "n_roots",
  "count"
)

h2_condition <- with(
  diagnostics,
  icc_at_zero > tau &
    icc_at_infinity < tau
)

h2_all_conditions <- all(
  h2_condition
)

h2_all_unique_positive_roots <- all(
  diagnostics$n_roots == 1L &
    is.finite(diagnostics$radius_1) &
    diagnostics$radius_1 > 0 &
    is.na(diagnostics$radius_2)
)

h2_min_zero_margin <- min(
  diagnostics$icc_at_zero - tau
)

h2_min_infinity_margin <- min(
  tau - diagnostics$icc_at_infinity
)

h6_trajectory <- trajectory[
  trajectory$step > 1L,
  ,
  drop = FALSE
]

h6_trajectory$z_observed <- with(
  h6_trajectory,
  radius / sqrt(group_offset_ss_before)
)

h6_trajectory$z_formula <- with(
  h6_trajectory,
  normalized_frontier_radius(
    n = group_size_before,
    g = state$n_groups,
    tau = tau,
    alignment = alignment_fixed
  )
)

h6_max_z_error <- max(
  abs(
    h6_trajectory$z_observed -
      h6_trajectory$z_formula
  )
)

h6_trajectory$a_multiplier_formula <- with(
  h6_trajectory,
  1 +
    2 * alignment_fixed * z_formula /
      (group_size_before + 1) +
    z_formula^2 /
      (group_size_before + 1)^2
)

h6_trajectory$group_offset_ss_after_formula <- with(
  h6_trajectory,
  group_offset_ss_before *
    a_multiplier_formula
)

h6_max_a_relative_error <- max(
  abs(
    h6_trajectory$group_offset_ss_after -
      h6_trajectory$group_offset_ss_after_formula
  ) /
    pmax(
      abs(h6_trajectory$group_offset_ss_after),
      .Machine$double.eps
    )
)

history_alignments <- c(
  -1,
  0,
  1
)

evaluation_alignments <- c(
  -0.5,
  0,
  0.5
)

history_states <- lapply(
  history_alignments,
  function(history_alignment) {
    gotham_sequence(
      state = state,
      steps = 10,
      threshold = tau,
      alignment = history_alignment
    )$final_state
  }
)

h6_history_check <- do.call(
  rbind,
  lapply(
    seq_along(history_states),
    function(history_index) {
      history_state <- history_states[[history_index]]

      do.call(
        rbind,
        lapply(
          evaluation_alignments,
          function(evaluation_alignment) {
            frontier <- gotham_frontier(
              state = history_state,
              threshold = tau,
              alignment = evaluation_alignment
            )

            data.frame(
              history_alignment =
                history_alignments[[history_index]],
              evaluation_alignment =
                evaluation_alignment,
              group_size =
                history_state$group_size,
              group_offset_ss =
                history_state$group_offset_ss,
              normalized_radius =
                frontier$frontier$radius_1[[1L]] /
                sqrt(
                  history_state$group_offset_ss
                ),
              formula_radius =
                normalized_frontier_radius(
                  n = history_state$group_size,
                  g = history_state$n_groups,
                  tau = tau,
                  alignment =
                    evaluation_alignment
                ),
              stringsAsFactors = FALSE
            )
          }
        )
      )
    }
  )
)

h6_history_check$error <- with(
  h6_history_check,
  normalized_radius - formula_radius
)

h6_max_history_error <- max(
  abs(
    h6_history_check$error
  )
)

h6_history_spread <- aggregate(
  normalized_radius ~ evaluation_alignment,
  data = h6_history_check,
  FUN = function(x) max(x) - min(x)
)

names(h6_history_spread)[[2L]] <-
  "normalized_radius_spread"

h6_max_history_spread <- max(
  h6_history_spread$normalized_radius_spread
)

h6_asymptotic <- data.frame(
  alignment = alignment_grid,
  z_infinity = normalized_frontier_radius_limit(
    g = state$n_groups,
    tau = tau,
    alignment = alignment_grid
  ),
  stringsAsFactors = FALSE
)

h6_asymptotic$group_offset_power <- with(
  h6_asymptotic,
  2 * alignment * z_infinity
)

h6_asymptotic$radius_power <- with(
  h6_asymptotic,
  alignment * z_infinity
)

threshold_state <- gotham_sequence(
  state = state,
  steps = 1,
  threshold = tau,
  alignment = 0
)$final_state

h7_periods <- list(
  zero = 0,
  plus_minus_0_25 = c(-0.25, 0.25),
  plus_minus_0_50 = c(-0.50, 0.50),
  plus_minus_1_00 = c(-1.00, 1.00)
)

h7_steps <- 5000L

h7_results <- do.call(
  rbind,
  lapply(
    names(h7_periods),
    function(schedule_name) {
      period <- h7_periods[[schedule_name]]

      alignments <- rep(
        period,
        length.out = h7_steps
      )

      path <- run_normalized_recurrence(
        initial_A =
          threshold_state$group_offset_ss,
        initial_n =
          threshold_state$group_size,
        alignment = alignments,
        g = threshold_state$n_groups,
        tau = tau
      )

      predicted_exponent <- mean(
        asymptotic_alignment_drift(
          g = threshold_state$n_groups,
          tau = tau,
          alignment = period
        )
      )

      late <- path$step >= floor(
        0.5 * h7_steps
      )

      estimated_exponent <- unname(
        coef(
          lm(
            log(group_offset_ss) ~
              log(group_size),
            data = path[late, , drop = FALSE]
          )
        )[[2L]]
      )

      data.frame(
        schedule = schedule_name,
        raw_mean_alignment = mean(period),
        predicted_exponent = predicted_exponent,
        estimated_late_exponent = estimated_exponent,
        final_group_offset_ratio =
          tail(path$group_offset_ss, 1L) /
          path$group_offset_ss[[1L]],
        stringsAsFactors = FALSE
      )
    }
  )
)

h7_alt_alignment <- rep(
  c(-0.5, 0.5),
  50L
)

h7_alt_sequence <- gotham_sequence(
  state = threshold_state,
  steps = length(h7_alt_alignment),
  threshold = tau,
  alignment = h7_alt_alignment
)

h7_alt_recurrence <- run_normalized_recurrence(
  initial_A =
    threshold_state$group_offset_ss,
  initial_n =
    threshold_state$group_size,
  alignment = h7_alt_alignment,
  g = threshold_state$n_groups,
  tau = tau
)

h7_alt_package_A <- c(
  threshold_state$group_offset_ss,
  h7_alt_sequence$trajectory$group_offset_ss_after
)

h7_alt_max_relative_error <- max(
  abs(
    h7_alt_package_A -
      h7_alt_recurrence$group_offset_ss
  ) /
    pmax(
      abs(h7_alt_package_A),
      .Machine$double.eps
    )
)

h7_a_infinity <- tau *
  (state$n_groups - 1) /
  (
    state$n_groups *
      (1 - tau)
  )

h7_symmetric_pair <- data.frame(
  magnitude = c(
    0.25,
    0.50,
    1.00
  ),
  pair_mean_alignment = 0,
  predicted_pair_average_exponent =
    2 * c(
      0.25,
      0.50,
      1.00
    )^2 /
    h7_a_infinity,
  stringsAsFactors = FALSE
)

h8_magnitudes <- c(
  0.25,
  0.50,
  1.00
)

h8_asymptotic_weights <- data.frame(
  magnitude = h8_magnitudes,
  positive_share = vapply(
    h8_magnitudes,
    function(magnitude) {
      neutral_positive_share_asymptotic(
        g = state$n_groups,
        tau = tau,
        negative_alignment = -magnitude,
        positive_alignment = magnitude
      )
    },
    numeric(1)
  ),
  stringsAsFactors = FALSE
)

h8_asymptotic_weights$negative_share <-
  1 -
  h8_asymptotic_weights$positive_share

h8_asymptotic_weights$weighted_drift <- vapply(
  h8_magnitudes,
  function(magnitude) {
    p_positive <- neutral_positive_share_asymptotic(
      g = state$n_groups,
      tau = tau,
      negative_alignment = -magnitude,
      positive_alignment = magnitude
    )

    h_positive <- asymptotic_alignment_drift(
      g = state$n_groups,
      tau = tau,
      alignment = magnitude
    )

    h_negative <- asymptotic_alignment_drift(
      g = state$n_groups,
      tau = tau,
      alignment = -magnitude
    )

    p_positive * h_positive +
      (1 - p_positive) *
      h_negative
  },
  numeric(1)
)

h8_finite_n_grid <- c(
  4L,
  10L,
  50L,
  100L,
  500L
)

h8_finite_n_weights <- data.frame(
  group_size = h8_finite_n_grid,
  positive_share = vapply(
    h8_finite_n_grid,
    function(n_value) {
      neutral_positive_share_finite_n(
        n = n_value,
        g = state$n_groups,
        tau = tau,
        negative_alignment = -0.5,
        positive_alignment = 0.5
      )
    },
    numeric(1)
  ),
  stringsAsFactors = FALSE
)

h8_finite_n_weights$negative_share <-
  1 -
  h8_finite_n_weights$positive_share

h8_finite_n_weights$asymptotic_positive_share <-
  neutral_positive_share_asymptotic(
    g = state$n_groups,
    tau = tau,
    negative_alignment = -0.5,
    positive_alignment = 0.5
  )

h8_greedy_results <- do.call(
  rbind,
  lapply(
    h8_magnitudes,
    function(magnitude) {
      path <- run_greedy_neutral_controller(
        initial_A =
          threshold_state$group_offset_ss,
        initial_n =
          threshold_state$group_size,
        steps = 5000L,
        g = threshold_state$n_groups,
        tau = tau,
        negative_alignment = -magnitude,
        positive_alignment = magnitude
      )

      selected <- path$alignment[
        !is.na(path$alignment)
      ]

      data.frame(
        magnitude = magnitude,
        asymptotic_positive_share =
          neutral_positive_share_asymptotic(
            g = threshold_state$n_groups,
            tau = tau,
            negative_alignment = -magnitude,
            positive_alignment = magnitude
          ),
        observed_positive_share =
          mean(
            selected > 0
          ),
        final_group_offset_ratio =
          tail(
            path$group_offset_ss,
            1L
          ) /
          path$group_offset_ss[[1L]],
        max_absolute_log_deviation =
          max(
            abs(
              path$log_ratio_to_initial
            )
          ),
        stringsAsFactors = FALSE
      )
    }
  )
)

h8_package_controller <- run_greedy_neutral_controller(
  initial_A =
    threshold_state$group_offset_ss,
  initial_n =
    threshold_state$group_size,
  steps = 100L,
  g = threshold_state$n_groups,
  tau = tau,
  negative_alignment = -0.5,
  positive_alignment = 0.5
)

h8_package_alignments <-
  h8_package_controller$alignment[
    !is.na(
      h8_package_controller$alignment
    )
  ]

h8_package_sequence <- gotham_sequence(
  state = threshold_state,
  steps = length(
    h8_package_alignments
  ),
  threshold = tau,
  alignment = h8_package_alignments
)

h8_package_A <- c(
  threshold_state$group_offset_ss,
  h8_package_sequence$trajectory$group_offset_ss_after
)

h8_controller_max_relative_error <- max(
  abs(
    h8_package_A -
      h8_package_controller$group_offset_ss
  ) /
    pmax(
      abs(h8_package_A),
      .Machine$double.eps
    )
)

h9_n_grid <- c(
  4L,
  5L,
  10L,
  25L,
  50L,
  100L,
  500L,
  1000L
)

h9_neutral_grid <- data.frame(
  group_size = h9_n_grid,
  neutral_alignment = vapply(
    h9_n_grid,
    function(n_value) {
      exact_structural_neutral_alignment(
        n = n_value,
        g = state$n_groups,
        tau = tau
      )
    },
    numeric(1)
  ),
  neutral_normalized_radius = vapply(
    h9_n_grid,
    function(n_value) {
      exact_structural_neutral_normalized_radius(
        n = n_value,
        g = state$n_groups,
        tau = tau
      )
    },
    numeric(1)
  ),
  stringsAsFactors = FALSE
)

h9_z_zero_limit <- sqrt(
  state$n_groups *
    (1 - tau) /
    (
      tau *
        (state$n_groups - 1)
    )
)

h9_neutral_grid$scaled_alignment <-
  h9_neutral_grid$group_size *
  h9_neutral_grid$neutral_alignment

h9_neutral_grid$asymptotic_scaled_alignment <-
  -0.5 *
  h9_z_zero_limit

h9_neutral_grid$asymptotic_normalized_radius <-
  h9_z_zero_limit

h9_steps <- 200L

h9_alignments <- vapply(
  seq_len(h9_steps),
  function(i) {
    n_value <-
      threshold_state$group_size +
      i -
      1L

    exact_structural_neutral_alignment(
      n = n_value,
      g = threshold_state$n_groups,
      tau = tau
    )
  },
  numeric(1)
)

h9_sequence <- gotham_sequence(
  state = threshold_state,
  steps = h9_steps,
  threshold = tau,
  alignment = h9_alignments
)

h9_A <- c(
  threshold_state$group_offset_ss,
  h9_sequence$trajectory$group_offset_ss_after
)

h9_max_relative_A_drift <- max(
  abs(
    h9_A /
      threshold_state$group_offset_ss -
      1
  )
)

h9_max_icc_error <- max(
  abs(
    h9_sequence$trajectory$icc_after -
      tau
  )
)

h9_observed_z <- with(
  h9_sequence$trajectory,
  radius /
    sqrt(
      group_offset_ss_before
    )
)

h9_formula_z <- vapply(
  seq_len(h9_steps),
  function(i) {
    n_value <-
      threshold_state$group_size +
      i -
      1L

    exact_structural_neutral_normalized_radius(
      n = n_value,
      g = threshold_state$n_groups,
      tau = tau
    )
  },
  numeric(1)
)

h9_max_z_error <- max(
  abs(
    h9_observed_z -
      h9_formula_z
  )
)

h9_selected_steps <- c(
  1L,
  2L,
  10L,
  50L,
  100L,
  200L
)

h9_sequence_table <- data.frame(
  step = h9_selected_steps,
  group_size_before =
    h9_sequence$trajectory$group_size_before[
      h9_selected_steps
    ],
  alignment =
    h9_sequence$trajectory$alignment_requested[
      h9_selected_steps
    ],
  radius =
    h9_sequence$trajectory$radius[
      h9_selected_steps
    ],
  group_offset_ss_before =
    h9_sequence$trajectory$group_offset_ss_before[
      h9_selected_steps
    ],
  group_offset_ss_after =
    h9_sequence$trajectory$group_offset_ss_after[
      h9_selected_steps
    ],
  icc_after =
    h9_sequence$trajectory$icc_after[
      h9_selected_steps
    ],
  stringsAsFactors = FALSE
)

find_endpoint_bracket <- function(
    summary,
    column,
    target = 1) {
  ordered <- summary[
    order(summary$alignment),
    c("alignment", column),
    drop = FALSE
  ]

  delta <- ordered[[column]] - target

  for (i in seq_len(nrow(ordered) - 1L)) {
    if (delta[[i]] == 0) {
      return(
        c(
          lower = ordered$alignment[[i]],
          upper = ordered$alignment[[i]]
        )
      )
    }

    if (
      is.finite(delta[[i]]) &&
        is.finite(delta[[i + 1L]]) &&
        sign(delta[[i]]) != sign(delta[[i + 1L]])
    ) {
      return(
        c(
          lower = ordered$alignment[[i]],
          upper = ordered$alignment[[i + 1L]]
        )
      )
    }
  }

  c(
    lower = NA_real_,
    upper = NA_real_
  )
}

offset_neutral_bracket <- find_endpoint_bracket(
  endpoint,
  "group_offset_ss_ratio"
)

radius_neutral_bracket <- find_endpoint_bracket(
  endpoint,
  "radius_ratio"
)

h10_same_mean_distributions <- list(
  narrow_negative = list(
    values = c(-0.15, -0.05),
    prob = c(0.50, 0.50)
  ),
  moderate_mixed = list(
    values = c(-0.50, 0.50),
    prob = c(0.60, 0.40)
  ),
  wide_mixed = list(
    values = c(-1.00, 1.00),
    prob = c(0.55, 0.45)
  )
)

h10_iid_review <- run_gotham_iid_alignment_review(
  distributions = h10_same_mean_distributions,
  initial_n = threshold_state$group_size,
  steps = 5000L,
  reps = 400L,
  g = threshold_state$n_groups,
  tau = tau,
  seed = 20261005L
)

h10_zero_mean_examples <- data.frame(
  scenario = c(
    "degenerate_zero",
    "symmetric_pm_0_5",
    "skewed_zero_mean"
  ),
  mean_alignment = c(
    0,
    0,
    0
  ),
  asymptotic_drift = c(
    gotham_asymptotic_drift(
      g = state$n_groups,
      tau = tau,
      alignment = 0
    ),
    mean(
      gotham_asymptotic_drift(
        g = state$n_groups,
        tau = tau,
        alignment = c(-0.5, 0.5)
      )
    ),
    sum(
      c(1 / 3, 2 / 3) *
        gotham_asymptotic_drift(
          g = state$n_groups,
          tau = tau,
          alignment = c(-1, 0.5)
        )
    )
  ),
  stringsAsFactors = FALSE
)

h10_neutral_positive_share <-
  neutral_positive_share_asymptotic(
    g = state$n_groups,
    tau = tau,
    negative_alignment = -0.5,
    positive_alignment = 0.5
  )

h10_neutral_distribution <- list(
  neutral_pm_0_5 = list(
    values = c(-0.5, 0.5),
    prob = c(
      1 - h10_neutral_positive_share,
      h10_neutral_positive_share
    )
  )
)

h10_neutral_review <- run_gotham_iid_alignment_review(
  distributions = h10_neutral_distribution,
  initial_n = threshold_state$group_size,
  steps = 5000L,
  reps = 400L,
  g = threshold_state$n_groups,
  tau = tau,
  seed = 20261007L
)

h11_markov_review <- run_gotham_markov_alignment_review(
  negative_alignment = -0.5,
  positive_alignment = 0.5,
  stationary_positive_share = 0.4,
  rho = c(-0.5, 0, 0.8),
  initial_n = threshold_state$group_size,
  steps = 5000L,
  reps = 400L,
  g = threshold_state$n_groups,
  tau = tau,
  seed = 20261006L
)

hypotheses <- data.frame(
  ID = c(
    "H1",
    "H2",
    "H3",
    "H4",
    "H5",
    "H6",
    "H7",
    "H8",
    "H9",
    "H10",
    "H11"
  ),
  Candidate = c(
    "Zero-radius headroom is path-invariant",
    "Threshold-hugging centered rays have one positive frontier crossing under a simple sufficient condition",
    "Orthogonality is not structural neutrality",
    "Structural neutrality is finite-n opposing and asymptotically orthogonal",
    "Radius and structural neutrality can differ at finite horizons but share the same asymptotic fixed-alignment boundary",
    "Threshold-state dynamics admit a scale-free normalized recurrence",
    "Mean-zero variable alignment is not generally Gotham-neutral",
    "Gotham-neutral mixed policies require asymmetric directional weighting",
    "The exact one-step structurally neutral alignment is slightly opposing",
    "IID stochastic Gotham drift is governed by E[h(C)], not E[C]",
    "Stationary dependence preserves the first-order drift exponent but changes fluctuations"
  ),
  Current_status = c(
    "Exact identity derived and numerically verified",
    "Exact sufficient condition derived and numerically verified",
    "Exact consequence of the update identity",
    "Reframed from H6/H9: exact one-step c_n* < 0; asymptotic fixed-alignment boundary c = 0",
    "Reframed: finite-horizon endpoint distinction is exact; no second asymptotic boundary",
    "Exact normalized recurrence derived and numerically verified",
    "Exact cumulative-drift criterion derived; symmetric alternation expands",
    "Exact asymptotic weighting rule and finite-n log-neutral policy derived and verified",
    "Exact neutral alignment and radius derived; package verification added",
    "IID almost-sure drift theorem derived; convexity corollary and simulations added",
    "Stationary-ergodic first-order result derived; Markov dependence simulation added"
  ),
  stringsAsFactors = FALSE
)
```

# Purpose

This is an internal research notebook for the statistical Gotham Loops framework. It is intentionally exploratory rather than package-facing documentation. The goal is to connect reproducible plots and numerical diagnostics to candidate mathematical statements, then decide which statements are worth deeper simulation, derivation, or formal proof.

The current notebook uses the canonical balanced three-group example, a fixed ICC threshold of 0.80, 50 sequential updates, and fixed alignments from -1 to 1 in increments of 0.25. These values are illustrative rather than universal.

# Hypothesis Registry

```{r hypothesis-registry}
knitr::kable(
  hypotheses,
  row.names = FALSE
)
```

# Current Sequence Geometry

## Relative Frontier Radius

```{r radius-trajectories, fig.width=7, fig.height=5}
plot_gotham_sequence_grid(
  review,
  metric = "radius",
  log_y = TRUE,
  relative = TRUE
)
```

## Relative Between-Group Geometry

```{r group-offset-trajectories, fig.width=7, fig.height=5}
plot_gotham_sequence_grid(
  review,
  metric = "group_offset_ss",
  log_y = TRUE,
  relative = TRUE
)
```

## 50-Step Endpoint

```{r endpoint-plot, fig.width=7, fig.height=5}
plot_gotham_sequence_grid_endpoint(
  review,
  metric = "group_offset_ss_ratio"
)
```

The 50-step endpoint ratios are:

```{r endpoint-table}
knitr::kable(
  endpoint[
    ,
    c(
      "alignment",
      "radius_ratio",
      "group_offset_ss_ratio",
      "ss_within_ratio",
      "min_icc_at_zero",
      "any_two_roots"
    )
  ],
  digits = 5,
  row.names = FALSE
)
```

# H1. Zero-Radius Headroom Is Path-Invariant

## Numerical observation

At a common step, all fixed-alignment trajectories return the same `icc_at_zero` value even though their group-separation and within-group sums of squares are very different.

The largest observed alignment-to-alignment spread in `icc_at_zero` over the 50-step review is:

```{r zero-spread}
max(
  zero_spread$spread
)
```

```{r zero-headroom-plot, fig.width=7, fig.height=4}
zero_one_path <- diagnostics[
  diagnostics$alignment == diagnostics$alignment[[1L]],
  ,
  drop = FALSE
]

plot(
  zero_one_path$step,
  zero_one_path$icc_at_zero,
  type = "l",
  lwd = 2,
  xlab = "Step",
  ylab = "ICC at zero-radius proposal",
  main = "Zero-radius ICC headroom"
)

abline(
  h = tau,
  lty = 2
)
```

## Exact identity

Let the current balanced state have common group size \(n\) and current ICC \(\rho\). Define

\[
q = \frac{MS_B}{MS_W}.
\]

The one-way ANOVA ICC identity gives

\[
\rho
=
\frac{q-1}{q+n-1},
\]

so

\[
q
=
\frac{1+\rho(n-1)}{1-\rho}.
\]

Now add one observation to every group exactly at that group's current mean. The group means and grand mean do not change. Therefore the within-group sum of squares is unchanged, while the between-group sum of squares is multiplied by \((n+1)/n\). Because the within-group degrees of freedom increase from \(G(n-1)\) to \(Gn\),

\[
MS_B^{+}
=
\frac{n+1}{n}MS_B
\]

and

\[
MS_W^{+}
=
\frac{n-1}{n}MS_W.
\]

The new common group size is \(n+1\), so the updated ICC is

\[
\rho_0
=
\frac{
MS_B^{+}-MS_W^{+}
}{
MS_B^{+}+nMS_W^{+}
}.
\]

Substituting \(q=MS_B/MS_W\) gives

\[
\rho_0
=
\frac{
(n+1)q-(n-1)
}{
(n+1)q+n(n-1)
}.
\]

Substituting the expression for \(q\) and simplifying yields

\[
\boxed{
\rho_0(n,\rho)
=
\frac{
n^2\rho+n\rho-2\rho+2
}{
n^2+n\rho-\rho+1
}
}.
\]

The zero-radius headroom is therefore

\[
\boxed{
\rho_0(n,\rho)-\rho
=
\frac{
(1-\rho)(n\rho-\rho+2)
}{
n^2+n\rho-\rho+1
}
}.
\]

For \(0 \le \rho < 1\), this quantity is strictly positive. Thus adding one observation exactly at every current group mean always increases the ANOVA ICC under the present balanced setup.

The identity contains neither \(G\) nor the separate values of \(SS_B\) and \(SS_W\). Conditional on the current group size and current ICC, the zero-radius ICC is therefore independent of the alignment path and of the absolute scale of the grouped data.

For fixed \(0 < \rho < 1\),

\[
\rho_0(n,\rho)-\rho
=
\frac{\rho(1-\rho)}{n}
+
O(n^{-2}),
\]

so the headroom decays at rate \(1/n\). At \(\rho=0\), the leading rate is \(O(n^{-2})\).

## Numerical verification

The first step of the canonical sequence does **not** begin at the target threshold: its current ICC is approximately 0.89655. This is why evaluating the formula with \(\tau=0.80\) for all 450 diagnostics produced the earlier maximum discrepancy of about 0.06776.

Using each row's actual `icc_before`, the maximum absolute discrepancy between the identity and `gotham_frontier()` across all 450 step-by-alignment diagnostics is:

```{r zero-formula-error}
zero_formula_max_error
```

After the first update, every trajectory begins each subsequent step at \(\rho=\tau=0.80\). Restricting the threshold-formula check to steps 2 through 50 gives maximum absolute discrepancy:

```{r zero-threshold-formula-error}
zero_threshold_formula_max_error
```

For the canonical first step, \(n=3\) and \(\rho=26/29\), giving:

```{r zero-first-step-check}
zero_icc_formula(
  n = 3,
  rho = 26 / 29
)
```

At step 50, the pre-update common group size is \(n=52\) and \(\rho=0.80\), giving:

```{r zero-step-50-check}
zero_icc_formula(
  n = 52,
  rho = 0.80
)
```

which matches the observed minimum `icc_at_zero` of approximately 0.8031175.

**Status:** H1 is proved exactly under the current balanced one-way ANOVA ICC formulation and verified numerically against the package calculations.

# H2. Why No Two-Root Behavior Appeared

Across the 50-step alignment grid, the observed frontier-root counts are:

```{r root-count-table}
knitr::kable(
  root_counts,
  row.names = FALSE
)
```

No two-root behavior appeared.

## Exact sufficient condition

For a centered proposal, write

\[
A = ||\mathbf{a}||^2
\]

for the current group-offset sum of squares and \(W=SS_W\) for the current within-group sum of squares. At target ICC \(\tau\), define \(C_\tau\) so that the frontier condition is

\[
SS_B' = C_\tau SS_W'.
\]

For current common group size \(n\),

\[
C_\tau
=
\frac{
(G-1)(1+n\tau)
}{
Gn(1-\tau)
}.
\]

Along a centered ray with radius \(r\) and alignment \(c\),

\[
SS_B'
=
(n+1)A
+
2\sqrt{A}\,cr
+
\frac{r^2}{n+1}
\]

and

\[
SS_W'
=
W
+
\frac{n}{n+1}r^2.
\]

Therefore the threshold equation can be written

\[
f_c(r)
=
\alpha r^2
+
\beta r
+
\gamma
=
0,
\]

where

\[
\alpha
=
\frac{1-C_\tau n}{n+1},
\qquad
\beta
=
2\sqrt{A}\,c,
\qquad
\gamma
=
(n+1)A-C_\tau W.
\]

The sign of \(\gamma=f_c(0)\) records which side of the threshold contains the zero-radius proposal. In particular,

\[
\rho_0>\tau
\quad\Longleftrightarrow\quad
\gamma>0.
\]

As \(r\rightarrow\infty\), the ICC converges to

\[
\boxed{
\rho_\infty(n,G)
=
\frac{1}{
G+n(G-1)
}
}.
\]

Equivalently,

\[
\rho_\infty<\tau
\quad\Longleftrightarrow\quad
\alpha<0.
\]

Hence, whenever

\[
\boxed{
\rho_0>\tau>\rho_\infty
},
\]

we have

\[
\gamma>0
\quad\text{and}\quad
\alpha<0.
\]

The product of the two quadratic roots is

\[
\frac{\gamma}{\alpha}<0.
\]

The discriminant is also strictly positive because

\[
\beta^2-4\alpha\gamma
>
\beta^2
\ge 0.
\]

Thus the two real roots have opposite signs. There is exactly one positive frontier crossing for every alignment \(c\in[-1,1]\).

This proves the proposed sufficient condition.

## Threshold-hugging corollary

Suppose the current state itself is already on the threshold,

\[
\rho=\tau<1.
\]

H1 gives

\[
\rho_0(n,\tau)>\tau.
\]

Therefore the unique-positive-root condition reduces to

\[
\boxed{
\tau
>
\frac{1}{
G+n(G-1)
}
}.
\]

Because \(\rho_\infty(n,G)\) decreases as \(n\) increases, once this inequality holds at a threshold-hugging step it continues to hold at every later step.

Thus, for the present sequential process, two-root or re-entry behavior cannot suddenly emerge at a later step once the process is on the threshold and this inequality is satisfied.

This explains why extending the current 50-step run to hundreds of steps would not reveal two-root behavior under the same setup.

## Alignment monotonicity corollary

In this regime the positive root is

\[
r_\tau(c)
=
\frac{
-\beta-\sqrt{\beta^2-4\alpha\gamma}
}{
2\alpha
},
\]

because \(\alpha<0\). At that root,

\[
\frac{\partial f_c}{\partial r}
=
-\sqrt{\beta^2-4\alpha\gamma}
<
0.
\]

Implicit differentiation therefore gives

\[
\frac{dr_\tau}{dc}
=
\frac{
2\sqrt{A}\,r_\tau
}{
\sqrt{\beta^2-4\alpha\gamma}
}
>
0
\]

for nondegenerate \(A>0\) and \(r_\tau>0\).

So, for a fixed state satisfying the H2 conditions, the centered frontier radius is strictly increasing with alignment. This formally explains the smooth left-to-right ordering of the static Gotham frontier.

## Numerical verification

The candidate large-radius identity is verified exactly by the current frontier calculations:

```{r infinity-formula-error}
infinity_formula_max_error
```

Across all 450 step-by-alignment diagnostics, the sufficient inequalities

\[
\rho_0>\tau>\rho_\infty
\]

hold:

```{r h2-condition-check}
h2_all_conditions
```

The smallest observed zero-radius margin above the threshold is:

```{r h2-zero-margin}
h2_min_zero_margin
```

The smallest observed margin of the threshold above the infinite-radius limit is:

```{r h2-infinity-margin}
h2_min_infinity_margin
```

And every one of the 450 cases has exactly one finite positive root and no second non-negative root:

```{r h2-root-check}
h2_all_unique_positive_roots
```

**Status:** H2 is proved as a sufficient-condition result under the current centered balanced one-way ANOVA formulation. The general `gotham_frontier()` implementation should continue to allow zero, one, or two roots because states outside these conditions can have different frontier topology.

# H3. Orthogonality Is Not Structural Neutrality

For centered proposal radius \(r\), alignment \(c\), current group-offset norm \(\|\mathbf{a}\|\), and current group size \(n\), the updated group-offset sum of squares is

\[
A'
=
A
+
\frac{2\sqrt{A}\,r\,c}{n+1}
+
\frac{r^2}{(n+1)^2},
\]

where \(A=\|\mathbf{a}\|^2\).

At \(c=0\),

\[
A'
=
A
+
\frac{r^2}{(n+1)^2}
>
A
\]

whenever \(r>0\).

This explains why the orthogonal trajectory in the plot rises above 1 rather than remaining structurally flat. Zero alignment removes the linear cross-product term, but it does not remove the positive quadratic contribution from the proposal.

For the present 50-step run:

```{r orthogonal-endpoint}
knitr::kable(
  endpoint[
    endpoint$alignment == 0,
    c(
      "alignment",
      "radius_ratio",
      "group_offset_ss_ratio",
      "ss_within_ratio"
    ),
    drop = FALSE
  ],
  digits = 5,
  row.names = FALSE
)
```

**Status:** Exact consequence of the update identity. This is conceptually important because "orthogonal" and "neutral" should not be used interchangeably.

# H4. Structural Neutrality Is Slightly Opposing at Finite n and Orthogonal Asymptotically

## Original finite-horizon observation

The 50-step endpoint plot showed monotone ordering by fixed alignment. Negative alignment contracted between-group geometry strongly, whereas positive alignment expanded it strongly.

On the coarse alignment grid, the fixed alignment for which final group-offset SS changed from below to above its initial value was bracketed by:

```{r offset-bracket}
offset_neutral_bracket
```

This bracket lies between -0.25 and 0 in the present run.

The important clarification is that this quantity is a **finite-horizon fixed-alignment endpoint crossing**. It asks for a constant alignment \(c\) such that, after the chosen number of updates,

\[
\frac{A_T(c)}{A_0}=1.
\]

It is not the same object as the exact one-step neutral alignment at a particular group size.

## Reframing from H6 and H9

H6 resolves the long-run fixed-alignment question. For any fixed \(c\),

\[
c<0
\Rightarrow
A_n\rightarrow0,
\]

\[
c=0
\Rightarrow
A_n\rightarrow A_\infty
\quad\text{for some finite positive }A_\infty,
\]

and

\[
c>0
\Rightarrow
A_n\rightarrow\infty.
\]

Thus the **asymptotic fixed-alignment contraction/expansion boundary is exactly**

\[
\boxed{c=0}.
\]

H9 resolves a different, local question. At each finite group size there is an exact one-step structurally neutral alignment

\[
c_n^*<0
\]

that satisfies

\[
A_{n+1}=A_n.
\]

Moreover,

\[
c_n^*
=
-\frac{z_\infty(0)}{2n}
+
O(n^{-2}),
\]

so the exact one-step neutral direction approaches orthogonality from the opposing side.

The original H4 observation therefore contains three related but distinct objects:

1. **one-step structural neutrality:** the step-specific \(c_n^*\) from H9;
2. **finite-horizon fixed-alignment neutrality:** a constant \(c_{A,T}\) satisfying \(A_T(c_{A,T})=A_0\);
3. **asymptotic fixed-alignment boundary:** \(c=0\), from H6.

The original 50-step grid estimated the second object. H9 gives the first exactly, while H6 gives the third exactly.

**Status:** The original H4 hypothesis is no longer an unresolved asymptotic question. Its finite-\(n\), finite-horizon, and asymptotic meanings should be kept separate.

# H5. Radius and Structural Neutrality Differ at Finite Horizons, Not Asymptotically

## Original finite-horizon observation

The relative frontier-radius plot showed a second endpoint transition. At 50 steps, the fixed alignment where the final frontier radius changed from below to above its step-1 radius was bracketed by:

```{r radius-bracket}
radius_neutral_bracket
```

In the present grid this lies between 0 and 0.25, whereas the 50-step group-offset SS crossing lies between -0.25 and 0.

The \(c=0.25\) trajectory remains a useful transient example because the frontier radius initially falls and later grows beyond its step-1 value.

```{r radius-transition-table}
knitr::kable(
  endpoint[
    endpoint$alignment %in% c(0, 0.25, 0.5),
    c(
      "alignment",
      "first_radius",
      "min_radius",
      "last_radius",
      "radius_ratio",
      "group_offset_ss_ratio"
    ),
    drop = FALSE
  ],
  digits = 5,
  row.names = FALSE
)
```

## Exact finite-horizon relationship

H6 gives

\[
r_n(c)
=
z_n(c)\sqrt{A_n}.
\]

Therefore, between two group sizes \(n_0\) and \(n\),

\[
\boxed{
\frac{r_n(c)}{r_{n_0}(c)}
=
\frac{z_n(c)}{z_{n_0}(c)}
\sqrt{
\frac{A_n}{A_{n_0}}
}
}.
\]

This identity explains why finite-horizon radius neutrality and structural neutrality need not coincide.

If

\[
A_n=A_{n_0},
\]

then

\[
\frac{r_n}{r_{n_0}}
=
\frac{z_n}{z_{n_0}},
\]

which is not generally 1 at finite \(n\). Conversely, if

\[
r_n=r_{n_0},
\]

then

\[
\frac{A_n}{A_{n_0}}
=
\left(
\frac{z_{n_0}}{z_n}
\right)^2,
\]

which is not generally 1.

H9 gives an especially direct example. Under the exact step-specific structurally neutral policy, \(A_n\) is constant by construction, while

\[
r_n
=
z_n^*\sqrt{A}
\]

still changes as \(z_n^*\) converges toward \(z_\infty(0)\). Exact structural neutrality therefore does not imply a constant frontier radius at finite group size.

## Asymptotic reconciliation

For fixed alignment, H6 shows that

\[
z_n(c)\rightarrow z_\infty(c),
\]

where \(z_\infty(c)\) is finite and strictly positive in the current regime. The long-run qualitative behavior of \(r_n\) therefore follows the same fixed-alignment boundary as \(A_n\):

\[
c<0
\Rightarrow
r_n\rightarrow0,
\]

\[
c=0
\Rightarrow
r_n\rightarrow r_\infty
\quad\text{for some finite positive }r_\infty,
\]

and

\[
c>0
\Rightarrow
r_n\rightarrow\infty.
\]

Thus there is **not** a second asymptotic critical alignment for radius growth. The asymptotic fixed-alignment boundary is again

\[
\boxed{c=0}.
\]

The original H5 observation should therefore be retained as a real and mathematically explained **finite-horizon/transient distinction**, not as evidence for two different asymptotic neutral directions.

**Status:** H5 is reframed. Radius neutrality and structural neutrality can differ at finite horizons because the normalized frontier factor \(z_n(c)\) evolves with group size, but their fixed-alignment asymptotic qualitative boundary is the same.

# H6. Threshold-State Dynamics Admit a Scale-Free Recurrence

H6 can be derived exactly once the current state lies on the ICC threshold.

Let

\[
A_n
=
||\mathbf{a}_n||^2
\]

be the current group-offset sum of squares when each group contains \(n\) observations, and let

\[
W_n
=
SS_{W,n}.
\]

Suppose the current state satisfies

\[
ICC_n=\tau.
\]

## Threshold state fixes the within/between scale ratio

Write

\[
q_n
=
\frac{MS_B}{MS_W}.
\]

From the one-way ANOVA ICC identity,

\[
\tau
=
\frac{
q_n-1
}{
q_n+n-1
},
\]

so

\[
q_n
=
\frac{
1+(n-1)\tau
}{
1-\tau
}.
\]

But

\[
MS_B
=
\frac{
nA_n
}{
G-1
}
\]

and

\[
MS_W
=
\frac{
W_n
}{
G(n-1)
}.
\]

Therefore

\[
\boxed{
\frac{W_n}{A_n}
=
\lambda_n
=
\frac{
nG(n-1)(1-\tau)
}{
(G-1)\{1+(n-1)\tau\}
}
}.
\]

Thus, on the threshold, \(W_n\) is proportional to \(A_n\), with a proportionality constant determined only by \(n\), \(G\), and \(\tau\).

## Normalize the frontier radius

Let the next centered proposal have radius \(r_n\) and alignment \(c\). Define the normalized radius

\[
z_n
=
\frac{
r_n
}{
\sqrt{A_n}
}.
\]

The next-step between-group and within-group sums of squares are

\[
SS_B'
=
(n+1)A_n
+
2\sqrt{A_n}\,c\,r_n
+
\frac{
r_n^2
}{
n+1
}
\]

and

\[
SS_W'
=
W_n
+
\frac{
n
}{
n+1
}r_n^2.
\]

For the updated state to lie on the same ICC threshold \(\tau\),

\[
SS_B'
=
C_{\tau,n}SS_W',
\]

where

\[
C_{\tau,n}
=
\frac{
(G-1)(1+n\tau)
}{
Gn(1-\tau)
}.
\]

Divide the frontier equation by \(A_n\), substitute \(r_n=z_n\sqrt{A_n}\), and substitute \(W_n/A_n=\lambda_n\). The absolute scale \(A_n\) disappears completely.

After simplification, the normalized frontier equation is

\[
-a_n z_n^2
+
2cz_n
+
\gamma_n
=
0,
\]

where

\[
\boxed{
a_n
=
\frac{
\tau\{G+n(G-1)\}-1
}{
G(n+1)(1-\tau)
}
}
\]

and

\[
\boxed{
\gamma_n
=
\frac{
\tau(n-1)+2
}{
\tau(n-1)+1
}
}.
\]

Under the H2 regime, \(a_n>0\). Therefore the unique positive normalized frontier radius is

\[
\boxed{
z_n(c)
=
\frac{
c+
\sqrt{
c^2+a_n\gamma_n
}
}{
a_n
}
}.
\]

This proves the scale-free part of H6:

\[
\boxed{
\frac{
r_n
}{
\sqrt{A_n}
}
=
z_n(c)
}
\]

depends only on \(n\), \(G\), \(\tau\), and \(c\), not on the absolute scale of \(A_n\) and not on the history that produced the threshold state.

## Exact multiplicative recurrence

The updated group-offset sum of squares is

\[
A_{n+1}
=
A_n
+
\frac{
2\sqrt{A_n}\,r_n c
}{
n+1
}
+
\frac{
r_n^2
}{
(n+1)^2
}.
\]

Using \(r_n=z_n\sqrt{A_n}\),

\[
\boxed{
A_{n+1}
=
A_n
\left[
1+
\frac{
2cz_n(c)
}{
n+1
}
+
\frac{
z_n(c)^2
}{
(n+1)^2
}
\right].
}
\]

Thus the sequential threshold-hugging process reduces to a one-dimensional multiplicative recurrence for \(A_n\) once the alignment sequence is specified.

For a fixed alignment \(c\), define

\[
m_n(c)
=
1+
\frac{
2cz_n(c)
}{
n+1
}
+
\frac{
z_n(c)^2
}{
(n+1)^2
}.
\]

Then

\[
\boxed{
A_{n+1}
=
A_nm_n(c).
}
\]

This is the exact normalized recurrence proposed in H6.

## Numerical verification against package trajectories

For steps 2 through 50 of all nine fixed-alignment trajectories, the maximum absolute discrepancy between

\[
r_n/\sqrt{A_n}
\]

from `gotham_sequence()` and the closed-form \(z_n(c)\) is:

```{r h6-z-error}
h6_max_z_error
```

The maximum relative discrepancy between the observed `group_offset_ss_after` and the exact multiplicative recurrence is:

```{r h6-a-error}
h6_max_a_relative_error
```

## History-invariance check

To check the path-invariance interpretation directly, three very different threshold states were generated after 10 steps using fixed history alignments \(-1\), \(0\), and \(1\). Their absolute group-offset scales differ substantially. Each state was then evaluated at common frontier alignments \(-0.5\), \(0\), and \(0.5\).

```{r h6-history-table}
knitr::kable(
  h6_history_check,
  digits = 8,
  row.names = FALSE
)
```

The maximum discrepancy from the closed-form normalized radius across these history/evaluation combinations is:

```{r h6-history-error}
h6_max_history_error
```

The largest spread in normalized radius across the three different histories at a common evaluation alignment is:

```{r h6-history-spread}
h6_max_history_spread
```

## Asymptotic consequence

For fixed \(G\), \(0<\tau<1\), and fixed alignment \(c\),

\[
a_n
\longrightarrow
a_\infty
=
\frac{
\tau(G-1)
}{
G(1-\tau)
}
\]

and

\[
\gamma_n
\longrightarrow
1.
\]

Therefore

\[
z_n(c)
\longrightarrow
z_\infty(c)
=
\frac{
c+\sqrt{c^2+a_\infty}
}{
a_\infty
}.
\]

The exact multiplicative factor satisfies

\[
m_n(c)
=
1+
\frac{
2cz_\infty(c)
}{
n
}
+
O(n^{-2}).
\]

Consequently,

\[
\log A_n
=
2cz_\infty(c)\log n
+
O(1),
\]

so, up to a positive multiplicative constant,

\[
\boxed{
A_n
\asymp
n^{\,2cz_\infty(c)}
}
\]

for \(c\ne0\). Because

\[
r_n
=
z_n(c)\sqrt{A_n},
\]

the frontier radius satisfies

\[
\boxed{
r_n
\asymp
n^{\,cz_\infty(c)}
}
\]

for \(c\ne0\).

At \(c=0\), the first-order \(1/n\) term vanishes. Instead,

\[
m_n(0)
=
1+
O(n^{-2}),
\]

and the infinite product of these factors converges. Therefore \(A_n\) increases by finite total multiplication and converges to a finite positive limit; \(r_n\) also converges to a finite positive limit.

This gives three asymptotic regimes:

- \(c<0\): \(A_n\rightarrow0\) and \(r_n\rightarrow0\);
- \(c=0\): \(A_n\) and \(r_n\) approach finite positive limits;
- \(c>0\): \(A_n\rightarrow\infty\) and \(r_n\rightarrow\infty\).

For the canonical \(G=3\), \(\tau=0.80\) example, the limiting normalized radii and power-law exponents are:

```{r h6-asymptotic-table}
knitr::kable(
  h6_asymptotic,
  digits = 6,
  row.names = FALSE
)
```

## Consequence for H4 and H5

The recurrence clarifies the two apparent 50-step transition points.

The finite-horizon endpoint plot placed the group-offset crossing between \(-0.25\) and \(0\), while the frontier-radius crossing lay between \(0\) and \(0.25\). Those are finite-horizon comparisons against the original scale.

Asymptotically, however, the recurrence shows that the critical alignment for both qualitative regimes is exactly

\[
\boxed{c=0}.
\]

What differs at finite horizons is the amount of accumulated transient scaling before the asymptotic power law dominates.

Thus the current evidence does **not** support two distinct asymptotic critical alignments. Instead, it supports a shared asymptotic boundary at zero alignment plus distinct finite-horizon neutral points.

**Status:** H6 is derived exactly under the current threshold-hugging centered balanced one-way ANOVA formulation. The recurrence also resolves the main asymptotic part of H4 and qualifies the original interpretation of H5.

# H7. Mean-Zero Variable Alignment Is Not Generally Gotham-Neutral

H6 extends immediately to a variable alignment sequence

\[
c_n\in[-1,1].
\]

The exact recurrence becomes

\[
A_{n+1}
=
A_n m_n(c_n),
\]

with

\[
m_n(c_n)
=
1+
\frac{
2c_n z_n(c_n)
}{
n+1
}
+
\frac{
z_n(c_n)^2
}{
(n+1)^2
}.
\]

Therefore

\[
\boxed{
A_N
=
A_{n_0}
\prod_{n=n_0}^{N-1}
m_n(c_n)
}.
\]

The entire variable-alignment history enters through a product of known one-step multipliers.

## Asymptotic cumulative-drift criterion

Define

\[
h(c)
=
2c z_\infty(c).
\]

From H6,

\[
\log m_n(c_n)
=
\frac{
h(c_n)
}{
n
}
+
O(n^{-2}),
\]

uniformly for \(c_n\in[-1,1]\). Hence

\[
\boxed{
\log A_N
=
\log A_{n_0}
+
\sum_{n=n_0}^{N-1}
\frac{
h(c_n)
}{
n
}
+
O(1)
}.
\]

This gives a cumulative criterion:

- if the weighted drift sum diverges to \(+\infty\), then \(A_N\rightarrow\infty\);
- if it diverges to \(-\infty\), then \(A_N\rightarrow0\);
- if it converges, then \(A_N\) approaches a finite positive limit.

So the relevant long-run summary is not the ordinary mean of the alignments. It is the harmonic-time accumulation of the nonlinear drift \(h(c_n)\).

## Symmetric positive and negative alignments are not neutral

Let

\[
a_\infty
=
\frac{
\tau(G-1)
}{
G(1-\tau)
}.
\]

Recall

\[
z_\infty(c)
=
\frac{
c+\sqrt{c^2+a_\infty}
}{
a_\infty
}.
\]

For any \(c>0\),

\[
z_\infty(c)-z_\infty(-c)
=
\frac{
2c
}{
a_\infty
}.
\]

Therefore

\[
h(c)+h(-c)
=
2c z_\infty(c)
-
2c z_\infty(-c)
=
\frac{
4c^2
}{
a_\infty
}
>
0.
\]

Thus an exactly symmetric alternating sequence

\[
-c,+c,-c,+c,\ldots
\]

has raw mean alignment zero but positive asymptotic Gotham drift.

Its average power-law exponent for \(A_n\) is

\[
\boxed{
\bar h_{\pm c}
=
\frac{
2c^2
}{
a_\infty
}
>0.
}
\]

This is a strong asymmetry result: equal-magnitude reinforcement and opposition do not cancel in the long run.

For the canonical \(G=3\), \(\tau=0.80\) example:

```{r h7-symmetric-pair-table}
knitr::kable(
  h7_symmetric_pair,
  digits = 6,
  row.names = FALSE
)
```

All of these schedules have mean alignment exactly zero, yet all have positive predicted expansion exponents.

## Exact recurrence versus package sequence

For 100 threshold-hugging steps alternating between \(-0.5\) and \(+0.5\), the maximum relative discrepancy between the H6 recurrence and `gotham_sequence()` is:

```{r h7-package-recurrence-error}
h7_alt_max_relative_error
```

## Long-run numerical illustration

The exact normalized recurrence was propagated for 5,000 steps for several periodic schedules:

```{r h7-long-run-table}
knitr::kable(
  h7_results,
  digits = 6,
  row.names = FALSE
)
```

For the alternating schedules, the raw mean alignment is zero, but the fitted late-time log-log slope is positive and approaches the predicted exponent.

The all-zero schedule is different. Its first-order drift is exactly zero, leaving only the summable \(O(n^{-2})\) contribution identified in H6, so it approaches a finite positive limit.

## Interpretation

This adds an important distinction to the Gotham framework:

\[
\boxed{
\text{mean alignment } = 0
\quad\not\Rightarrow\quad
\text{long-run structural neutrality}.
}
\]

A sequence can contain equal amounts of positive and negative directional alignment and still expand the between-group geometry because reinforcement and opposition have asymmetric normalized frontier radii.

A genuinely neutral variable-alignment policy must balance the nonlinear drift \(h(c)\), not merely the raw alignments \(c\).

**Status:** H7 is an exact consequence of the H6 recurrence. The symmetric-alternation result is proved analytically and checked numerically against both the recurrence and `gotham_sequence()`.

# H8. Gotham-Neutral Mixed Policies Require Asymmetric Directional Weighting

H7 shows that an equal-frequency \(+c/-c\) schedule is not generally neutral. H8 asks what weighting would actually remove the long-run Gotham drift.

## General two-direction rule

Choose two available alignments

\[
c_-<0<c_+.
\]

Recall the asymptotic drift function

\[
h(c)
=
2c z_\infty(c).
\]

Because \(h(c_-)<0<h(c_+)\), a mixed policy can balance the two directions.

The relevant asymptotic weighting is a **harmonic-time share** because H7 showed that alignment contributions enter through \(1/n\). If the positive direction receives limiting harmonic-time share \(p_+\), asymptotic neutrality requires

\[
p_+h(c_+)
+
(1-p_+)h(c_-)
=
0.
\]

Solving gives

\[
\boxed{
p_+^*
=
\frac{
-h(c_-)
}{
h(c_+)-h(c_-)
}
}
\]

and

\[
\boxed{
p_-^*
=
1-p_+^*.
}
\]

For well-distributed periodic or stationary schedules, the harmonic-time share equals the ordinary long-run frequency. For arbitrarily block-structured schedules, raw frequency alone is not sufficient; the harmonic weighting remains the relevant criterion.

## Symmetric magnitudes

For the symmetric pair

\[
c_-=-c,
\qquad
c_+=c,
\qquad
c>0,
\]

let

\[
a_\infty
=
\frac{
\tau(G-1)
}{
G(1-\tau)
}
\]

and

\[
s
=
\sqrt{
c^2+a_\infty
}.
\]

Using the H7 drift identities,

\[
h(c)
=
\frac{
2c(c+s)
}{
a_\infty
}
\]

and

\[
h(-c)
=
-\frac{
2c(s-c)
}{
a_\infty
}.
\]

Therefore

\[
\boxed{
p_+^*
=
\frac{
s-c
}{
2s
}
=
\frac12
-
\frac{
c
}{
2\sqrt{c^2+a_\infty}
}
}
\]

and

\[
\boxed{
p_-^*
=
\frac{
s+c
}{
2s
}
=
\frac12
+
\frac{
c
}{
2\sqrt{c^2+a_\infty}
}
}.
\]

Thus, for every \(c>0\),

\[
p_+^*<\frac12<p_-^*.
\]

A neutral policy must use the opposing direction more often than the reinforcing direction.

For the canonical \(G=3,\tau=0.80\) example:

```{r h8-asymptotic-weight-table}
knitr::kable(
  h8_asymptotic_weights,
  digits = 6,
  row.names = FALSE
)
```

The residual weighted drift is numerical zero.

The directional imbalance becomes stronger as the available magnitude \(c\) grows because reinforcing moves have greater structural leverage.

## Exact finite-\(n\) log-neutral mixture

The asymptotic share is not the only possible neutrality rule. H6 gives the exact one-step multiplier

\[
m_n(c)
=
\frac{
A_{n+1}
}{
A_n
}.
\]

Suppose that at group size \(n\) a stochastic policy chooses \(c_+\) with probability \(p_n\) and \(c_-\) otherwise. Requiring the expected one-step change in \(\log A_n\) to equal zero gives

\[
p_n\log m_n(c_+)
+
(1-p_n)\log m_n(c_-)
=
0.
\]

Therefore the exact finite-\(n\) log-neutral probability is

\[
\boxed{
p_n^+
=
\frac{
-\log m_n(c_-)
}{
\log m_n(c_+)-\log m_n(c_-)
}
}.
\]

As \(n\rightarrow\infty\),

\[
p_n^+
\longrightarrow
p_+^*.
\]

For the symmetric \(\pm0.5\) policy:

```{r h8-finite-n-table}
knitr::kable(
  h8_finite_n_weights,
  digits = 6,
  row.names = FALSE
)
```

This provides an exact adaptive mixed policy if neutrality is defined on the log scale.

## Deterministic finite-horizon controller

A deterministic analogue can use the same exact multipliers. At each step, compare the two allowable moves and choose whichever makes the cumulative

\[
\log\left(
\frac{
A_n
}{
A_{\mathrm{start}}
}
\right)
\]

closest to zero.

This controller is not needed for the asymptotic proof, but it makes the neutrality principle concrete.

Over 5,000 threshold-hugging steps:

```{r h8-greedy-controller-table}
knitr::kable(
  h8_greedy_results,
  digits = 6,
  row.names = FALSE
)
```

The realized positive-direction frequencies approach the asymptotic weights, while the final group-offset SS remains extremely close to its starting value.

For the \(\pm0.5\) controller, a 100-step schedule generated from the exact recurrence was also passed back through `gotham_sequence()`. The maximum relative discrepancy in group-offset SS is:

```{r h8-package-controller-error}
h8_controller_max_relative_error
```

## Interpretation

H8 turns the qualitative asymmetry of H7 into an explicit control law.

For symmetric directions, a policy that uses reinforcement and opposition equally is structurally biased toward expansion. To remove that bias, the process must intentionally favor opposing moves.

The stronger the available directional magnitude, the more asymmetric the neutral weighting becomes.

This gives a precise Gotham distinction among:

- **directional balance:** equal positive and negative alignment frequency;
- **asymptotic structural balance:** zero harmonic-time nonlinear drift;
- **finite-\(n\) log balance:** zero expected one-step change in \(\log A_n\);
- **deterministic finite-horizon balance:** actively choosing directions to keep cumulative log geometry near its initial value.

These are not equivalent notions.

**Status:** H8 is derived exactly from H6 and H7. The asymptotic weighting rule and finite-\(n\) log-neutral rule are analytic; the deterministic controller is a constructive illustration using the exact recurrence.

# H9. The Exact One-Step Structurally Neutral Alignment Is Slightly Opposing

H3 showed that orthogonal alignment \(c=0\) is not structurally neutral: the positive quadratic term in the update increases \(A_n\). H9 asks whether there is a single centered alignment that leaves the group-offset geometry exactly unchanged at each threshold-hugging step.

The H6 recurrence is

\[
A_{n+1}
=
A_n
\left[
1+
\frac{
2cz_n(c)
}{
n+1
}
+
\frac{
z_n(c)^2
}{
(n+1)^2
}
\right].
\]

Exact one-step structural neutrality requires

\[
A_{n+1}=A_n.
\]

Because the frontier radius is positive, this reduces to

\[
\frac{
2cz_n
}{
n+1
}
+
\frac{
z_n^2
}{
(n+1)^2
}
=
0,
\]

hence

\[
\boxed{
z_n^*
=
-2c_n^*(n+1)
}.
\]

Therefore an exactly neutral single direction must satisfy

\[
c_n^*<0.
\]

It must be slightly opposing, not orthogonal.

## Closed-form neutral alignment

The normalized frontier equation from H6 is

\[
-a_n z_n^2
+
2cz_n
+
\gamma_n
=
0,
\]

where

\[
a_n
=
\frac{
\tau\{G+n(G-1)\}-1
}{
G(n+1)(1-\tau)
}
\]

and

\[
\gamma_n
=
\frac{
\tau(n-1)+2
}{
\tau(n-1)+1
}.
\]

Substitute

\[
z_n=-2c(n+1)
\]

into the frontier equation. After simplification,

\[
\gamma_n
=
4c^2(n+1)
\left[
a_n(n+1)+1
\right].
\]

Therefore the unique negative neutral alignment is

\[
\boxed{
c_n^*
=
-\frac12
\sqrt{
\frac{
G(1-\tau)\{\tau(n-1)+2\}
}{
(n+1)(G-1)(1+n\tau)\{\tau(n-1)+1\}
}
}
}.
\]

The associated normalized frontier radius is

\[
\boxed{
z_n^*
=
\sqrt{
\frac{
G(1-\tau)(n+1)\{\tau(n-1)+2\}
}{
(G-1)(1+n\tau)\{\tau(n-1)+1\}
}
}
}.
\]

This provides an exact deterministic rule that leaves the between-group geometry \(A_n\) unchanged after every update while still ending each step on the ICC frontier.

## Asymptotic behavior

Define the orthogonal limiting normalized radius from H6,

\[
z_\infty(0)
=
\sqrt{
\frac{
G(1-\tau)
}{
\tau(G-1)
}
}.
\]

Then

\[
\boxed{
c_n^*
=
-\frac{
z_\infty(0)
}{
2n
}
+
O(n^{-2})
}
\]

and

\[
\boxed{
z_n^*
\longrightarrow
z_\infty(0).
}
\]

So the exact finite-\(n\) neutral direction approaches orthogonality from the opposing side at rate \(1/n\).

This resolves the main finite-horizon question behind H4: the structurally neutral direction is not merely somewhere between a negative alignment and zero. It is available in closed form, is always slightly negative in the present regime, and converges to zero.

For the canonical \(G=3,\tau=0.80\) example:

```{r h9-neutral-grid-table}
knitr::kable(
  h9_neutral_grid,
  digits = 8,
  row.names = FALSE
)
```

The table shows both the exact neutral alignment and the convergence of \(n c_n^*\) and \(z_n^*\) to their asymptotic limits.

## Direct package verification

Starting from the canonical threshold state, `gotham_sequence()` was run for 200 steps using the exact step-specific neutral alignment \(c_n^*\).

Selected steps are:

```{r h9-sequence-table}
knitr::kable(
  h9_sequence_table,
  digits = 10,
  row.names = FALSE
)
```

The largest relative deviation of group-offset SS from its initial value across all 200 updates is:

```{r h9-A-drift}
h9_max_relative_A_drift
```

The largest discrepancy between observed normalized radius and the closed-form \(z_n^*\) is:

```{r h9-z-error}
h9_max_z_error
```

The largest post-update ICC discrepancy from \(\tau\) is:

```{r h9-icc-error}
h9_max_icc_error
```

All three should be at numerical tolerance.

## Interpretation

H9 gives a third notion of Gotham neutrality:

- H8 provides **mixed-policy neutrality**, achieved by asymmetrically weighting different directions;
- H9 provides **single-step deterministic neutrality**, achieved by choosing one slightly opposing alignment at each group size.

These are distinct control strategies that solve the same structural objective.

The result also explains why \(c=0\) was close to, but not exactly, neutral in the finite-horizon plots. Orthogonality removes the first-order directional cross-product, but the quadratic frontier term still increases group separation. A small opposing component is required to cancel that increase exactly.

As \(n\) grows, the required correction becomes smaller, which is why the asymptotic critical alignment from H6 is exactly zero.

**Status:** H9 is derived exactly from the H6 recurrence. The closed-form neutral alignment and normalized radius are verified directly against `gotham_sequence()`.


# H10. IID Stochastic Gotham Drift Is Governed by E[h(C)], Not E[C]

H7 treated deterministic variable-alignment sequences. Now let

\[
C_n
\]

be independent and identically distributed random alignments taking values in \([-1,1]\).

The exact H6 recurrence remains

\[
A_{n+1}
=
A_n m_n(C_n).
\]

Define the limiting drift function

\[
h(c)
=
2c z_\infty(c).
\]

## Uniform first-order expansion

Because \(c\in[-1,1]\), the normalized frontier radius converges uniformly:

\[
z_n(c)
=
z_\infty(c)
+
O(n^{-1}).
\]

The exact multiplier therefore satisfies

\[
\log m_n(c)
=
\frac{h(c)}{n}
+
O(n^{-2}),
\]

uniformly over the entire alignment interval.

Consequently,

\[
\log A_N
=
\log A_{n_0}
+
\sum_{n=n_0}^{N-1}
\frac{h(C_n)}{n}
+
R_N,
\]

where \(R_N\) converges absolutely to a finite random limit because the remainder is uniformly \(O(n^{-2})\).

Let

\[
\mu_h
=
E[h(C)].
\]

Then

\[
\sum_n
\frac{h(C_n)-\mu_h}{n}
\]

converges almost surely and in \(L^2\): the summands are independent, centered, bounded, and their variances are dominated by a constant times \(1/n^2\).

Thus

\[
\boxed{
\log A_N
-
\mu_h\log N
\longrightarrow
L
\quad\text{almost surely}
}
\]

for a finite random variable \(L\), up to the immaterial choice of the initial group-size index.

Equivalently,

\[
\boxed{
\frac{A_N}{N^{\mu_h}}
\longrightarrow
K
\in(0,\infty)
\quad\text{almost surely}.
}
\]

This gives the IID stochastic classification:

\[
\mu_h<0
\Rightarrow
A_N\rightarrow0
\quad\text{almost surely},
\]

\[
\mu_h=0
\Rightarrow
A_N\rightarrow K
\in(0,\infty)
\quad\text{almost surely},
\]

and

\[
\mu_h>0
\Rightarrow
A_N\rightarrow\infty
\quad\text{almost surely}.
\]

The stochastic analogue of the deterministic Gotham exponent is therefore

\[
\boxed{
\mu_h=E[h(C)],
}
\]

not \(E[C]\).

## A stronger convexity result

Write

\[
a_\infty
=
\frac{\tau(G-1)}{G(1-\tau)}
\]

and

\[
h(c)
=
\frac{
2c\{c+\sqrt{c^2+a_\infty}\}
}{
a_\infty
}.
\]

Let

\[
t
=
\frac{c}{\sqrt{c^2+a_\infty}},
\qquad
-1<t<1.
\]

Direct differentiation gives

\[
\boxed{
h''(c)
=
\frac{2}{a_\infty}
(1+t)^2(2-t)
>0.
}
\]

Thus \(h(c)\) is strictly convex.

Jensen's inequality then gives

\[
E[h(C)]
\ge
h(E[C]),
\]

with strict inequality for any non-degenerate alignment distribution.

In particular,

\[
\boxed{
E[C]=0
\ \text{and}\ 
\operatorname{Var}(C)>0
\quad\Longrightarrow\quad
E[h(C)]>0.
}
\]

So **every non-degenerate mean-zero IID alignment distribution is asymptotically expansive** in the present Gotham regime.

This substantially generalizes H7, which first demonstrated the phenomenon for symmetric deterministic alternation.

It also implies that any non-degenerate IID distribution satisfying

\[
E[h(C)]=0
\]

must have

\[
E[C]<0.
\]

Stochastic Gotham neutrality requires an opposing bias in ordinary alignment space.

For the canonical example, several mean-zero distributions illustrate the convexity result:

```{r h10-zero-mean-table}
knitr::kable(
  h10_zero_mean_examples,
  digits = 6,
  row.names = FALSE
)
```

## Same ordinary mean, different Gotham regimes

To demonstrate why \(E[C]\) is insufficient, consider three IID distributions that all have

\[
E[C]=-0.10.
\]

They differ only in how dispersed the alignment distribution is.

The exact recurrence was simulated for 5,000 steps and 400 independent realizations per distribution.

```{r h10-iid-table}
knitr::kable(
  h10_iid_review,
  digits = 6,
  row.names = FALSE
)
```

The narrow negative distribution has negative \(E[h(C)]\) and contracts. The more dispersed distributions have the **same ordinary mean alignment** but positive \(E[h(C)]\), so they expand.

The simulated mean log-growth exponents should closely track the exact finite-horizon expected exponent and move toward the asymptotic value \(E[h(C)]\).

## IID stochastic neutrality

For the symmetric support \(\{-0.5,+0.5\}\), H8 gives the asymptotically neutral positive-direction probability. Using that probability as an IID sampling rule gives \(E[h(C)]=0\).

```{r h10-neutral-table}
knitr::kable(
  h10_neutral_review,
  digits = 6,
  row.names = FALSE
)
```

The ordinary mean alignment of this neutral distribution is negative, while its Gotham drift is zero. H10 predicts convergence to a positive random finite limit rather than polynomial expansion or contraction.

**Status:** H10 gives an exact IID stochastic asymptotic theorem under the current centered threshold-hugging formulation. It identifies \(E[h(C)]\) as the governing exponent and shows by strict convexity that non-degenerate mean-zero IID alignment is necessarily expansive.

# H11. Dependence Preserves the First-Order Drift Exponent but Changes Fluctuations

The IID assumption is stronger than necessary for the first-order exponent.

Let

\[
H_n
=
h(C_n)
\]

and suppose \(\{C_n\}\) is stationary and ergodic with

\[
E|H_n|<\infty.
\]

Let

\[
\mu_h
=
E[H_n].
\]

The ergodic theorem gives

\[
S_N
=
\sum_{n=1}^{N}
(H_n-\mu_h)
=
o(N)
\quad\text{almost surely}.
\]

By summation by parts,

\[
\sum_{n=1}^{N}
\frac{H_n-\mu_h}{n}
=
\frac{S_N}{N}
+
\sum_{n=1}^{N-1}
\frac{S_n}{n(n+1)}.
\]

Since \(S_n=o(n)\),

\[
\frac{1}{\log N}
\sum_{n=1}^{N}
\frac{H_n-\mu_h}{n}
\longrightarrow
0
\quad\text{almost surely}.
\]

Combining this with the exact Gotham remainder gives

\[
\boxed{
\frac{\log A_N}{\log N}
\longrightarrow
\mu_h
=
E[h(C)]
\quad\text{almost surely}.
}
\]

Therefore a stationary ergodic alignment process with a fixed marginal distribution has the same **first-order polynomial exponent** regardless of its autocorrelation structure.

For

\[
\mu_h\neq0,
\]

the contraction-versus-expansion classification is unchanged by dependence.

The boundary case

\[
\mu_h=0
\]

is more delicate. Under IID sampling, H10 proves that the centered harmonic series converges and \(A_N\) approaches a finite positive random limit. Stationarity and ergodicity alone only imply

\[
\log A_N=o(\log N),
\]

so stronger short-range-dependence conditions are needed to obtain the same finite-limit conclusion.

## Two-state Markov illustration

To isolate the role of dependence, consider the same marginal distribution in every case:

\[
C_n\in\{-0.5,+0.5\},
\]

with stationary probability

\[
P(C_n=+0.5)=0.40.
\]

Thus every process has the same

\[
E[C]=-0.10
\]

and the same positive

\[
E[h(C)].
\]

Only the lag-one persistence parameter is changed.

The exact recurrence was propagated for 5,000 steps across 400 realizations for negatively correlated, IID, and positively persistent two-state chains.

```{r h11-markov-table}
knitr::kable(
  h11_markov_review,
  digits = 6,
  row.names = FALSE
)
```

The mean empirical exponent should remain close to the same finite-horizon and asymptotic predictions across all three dependence structures because the stationary marginal distribution is unchanged.

What dependence changes is the **finite-sample dispersion** around that exponent. Positive persistence clusters reinforcing or opposing moves into longer runs, while negative autocorrelation alternates directions more aggressively.

This separates two stochastic questions:

1. **first-order structural drift:** governed by the stationary marginal value \(E[h(C)]\);
2. **fluctuations and neutral-boundary behavior:** sensitive to dependence, autocorrelation, and mixing properties.

**Status:** H11 establishes the stationary-ergodic first-order exponent and introduces dependence through a controlled two-state Markov comparison. The next stochastic step, if needed, would be to characterize the variance/random-limit behavior under explicit short-range-dependence assumptions rather than adding more arbitrary simulation.

# Additional Plot Features Worth Following

The current plots suggest several secondary questions:

- **Strong asymmetry:** reinforcement produces far more expansion than equal-magnitude opposition produces contraction. The positive quadratic term in the group-offset update is a likely source of this asymmetry.
- **Smooth ordering:** the trajectories are smoothly ordered by alignment with no visible crossing in the current grid.
- **Transient behavior:** some mildly positive alignments first reduce the frontier radius and later increase it, suggesting finite-group-size dynamics before a possible long-run regime dominates.
- **Path-independent headroom:** despite enormous differences in `group_offset_ss` and `ss_within`, the zero-radius ICC is numerically identical across alignments at a common step.

These are observations from the current canonical example, not general results.

# Recommended Investigation Order

1. **H1 complete:** retain the exact zero-radius headroom identity.
2. **H2 complete:** retain the unique-positive-root sufficient condition and its threshold-hugging corollary.
3. **H6 complete:** retain the exact normalized recurrence and its fixed-alignment asymptotic regimes.
4. **H7 complete:** retain the variable-alignment cumulative-drift criterion and the result that symmetric \(+c/-c\) alignment is generically expansive.
5. **H8 complete:** retain the asymmetric neutral-weighting rule, finite-\(n\) log-neutral mixture, and deterministic mixed controller.
6. **H9 complete:** retain the exact per-step structurally neutral alignment and its direct package verification at numerical tolerance.
7. **H4/H5 reconciled:** distinguish one-step, finite-horizon, and asymptotic neutrality; retain the radius/structure distinction only as a finite-horizon phenomenon.
8. **H10 added:** derive and test IID stochastic alignment; classify long-run behavior by \(E[h(C)]\), not \(E[C]\).
9. **H11 added:** introduce stationary dependence and isolate its effect on first-order drift versus finite-sample fluctuations.
10. After verifying H10/H11 numerically, decide whether to deepen the stochastic theory at the neutral boundary or broaden the structural criterion beyond ICC.

The deterministic centered threshold-hugging model is now largely reducible to exact low-dimensional recurrences and control rules. Future simulation should target questions not already settled analytically.
